Classification of ordinal data in deep learning: an experimental study

### 2019-01-14 paper/ieee.tex
Paper

@@

	Ordinal classification problems are those where labels are ordered, and there are different inter-classes weights for each pair of classes. This kind of problem can be treated as a nominal classification problem, but when you do this, you are the discarding ordinal information. A better approach is to use specific methods that take into account this kind of information to improve the performance of the classification model. One way to use the ordinal information is to evaluate the model using an ordinal metric. Multiple metrics exist in the literature of machine learning and statistics \cite{cruz2014metrics, mehdiyev2016evaluating}. Kappa index is a well-known statistic coefficient defined by Cohen \cite{cohen1960coefficient} to measure inter-rater agreement on classifying elements into a set of categories. Later, Weighted Kappa (WK) is a modified version of the Kappa statistic calculated to allow assigning different weights to different levels of aggregation between two variables. Weighted Kappa loss was described in previous work~\cite{de2018weighted} and is a cost function that is based on the WK metric. These functions are indicated for problems where different inter-classes weights are assigned, and that work proved that using these functions improves the model performance and reduce the overfitting risk. This kind of problems is associated with ordinal data processing. Those weights are predefined and depend on the type of the chosen WK. In a linear penalization, the weight is proportional to the distance between the predicted and the real class. In the Quadratic Weighted Kappa (QWK) the penalization is proportional to the square of the distance. In this work, we will combine the QWK metric and the QWK loss.

@@

	This paper is organised as follows: in Section~\ref{sect:relatedwork}, we take a look at previous works related to this paper, in Section~\ref{sect:experiments} we describe the model, the experiments and the datasets used, in Section~\ref{sect:results} we present the results obtained and the statistical analysis and finally in Section~\ref{sect:conclusions} we expose the conclusions of this study.

@@

	\begin{figure}[!t]
		\centering
		\includegraphics[width=2.5in]{img/linkfunctions.png}
		\caption{Representation of link functions.}
		\label{fig:linkfunctions}
	\end{figure}

@@

	The model is evaluated with Quadratic Weighted Kappa metric (QWK)~\cite{ben2008comparison}. This evaluation measure gives a higher weight to the errors that are further from the correct class. Quadratic Weighted Kappa Loss is considered as loss function for the optimizer as it gives better performance for ordinal classification problems.

@@

	\subsection{Diabetic Retinopathy}
	Detailed results for the Diabetic Retinopathy dataset are presented in Table \ref{table:DRresults}.

@@

	The best mean WK value was obtained with the complementary log-log link function using a batch size of 10 and a learning rate of $10^{-3}$.

	\subsection{Adience}
	Results for the experiments made with the Adience dataset are shown in Table \ref{table:Adienceresults}.

@@

	The best mean WK value was obtained with the probit link function using a batch size of 15 and a learning rate of $10^{-2}$.

@@

	We denote by $WK_{i,j,k,l}(i=1, ..., 3; j = 1, ..., 3; k = 1, ..., 3)$ the value observed when the first factor is at the i-th level, the second at the j-th level and the third at k-th level. We assume that the two factors do not act independently and therefore there exists an interaction between them. In this case, the observations fit \ref{eq:anova3}.

	\begin{equation}
	\begin{aligned}
	\footnotesize
	WK_{i,j,k,l} = & \mu + L_i + P_j + B_k + LP_{i,j} + LB_{i,k} \\& + PB_{j,k} + LPB_{i,j,k} + \epsilon_{i,j,k,l}
	\label{eq:anova3}
	\end{aligned}
	\end{equation}

@@

	The analysis of variance table containing the sum of squares, degrees of freedom, mean square, test statistics and significance level represents the initial study in a compact form.

	\begin{table}[!t]
		\caption{ANOVA III for the analysis of the main factors in the design of a Convolutional Ordinal Neural Network for the Retinopathy dataset.}
		\label{table:ANOVARetinopathy}
		\centering
		\small
		\begin{tabular}{l|ccccc}
			\multicolumn{6}{c}{Response variable WK}\\\hline
			Source & S.S. & D.F. & M.S. & F-ratio & Sig.\\\hline
			Model & $37.860$ & $9$ & $4.207$ & $1562.840$ & $0.000$\\
			$L$ factor & $0.057$ & $2$ & $0.029$ & $10.646$ & $0.000$\\
			$P$ factor & $0.121$ & $2$ & $0.060$ & $22.468$ & $0.000$\\
			$LP$ factors & $0.057$ & $4$ & $0.014$ & $5.261$ & $0.001$\\
			Error & $0.339$ & $126$ & $0.003$ & & \\
			Total & $38.199$ & $135$ & & & 
		\end{tabular}
	\end{table}

	The results of the ANOVA III test for the Diabetic Retinopathy dataset are resumed in Table \ref{table:ANOVARetinopathy}. There are significative differences in the means depending on the link function and also depending on the learning rate for $\alpha=0.05$. Moreover, an interaction between the link function and the learning rate can be recognised ($\text{p-value} = 0.001$).

	As Table \ref{table:ANOVARetinopathy} showed that there exist significative differences between the means, we are now analysing those differences. A post-hoc multiple comparison test has been performed on the average WK obtained. An HSD Tukey's test was made under the null hypothesis that the variance of the error of the dependent variable is the same between groups. The best link function is the complementary log-log, though it doesn't have significative differences with the probit function ($\alpha=0.05$). There are differences between the complementary log-log and the probit function concerning the logit function.

@@

	\begin{table}[!t]
		\caption{ANOVA III for the analysis of the main factors in the design of a Convolutional Ordinal Neural Network for the Adience dataset.}
		\label{table:ANOVAAdience}
		\centering
		\small
		\begin{tabular}{l|ccccc}
			\multicolumn{6}{c}{Response variable WK}\\\hline
			Source & S.S. & D.F. & M.S. & F-ratio & Sig.\\\hline
			Model & $52.328$ & $15$ & $3.489$ & $1426.170$ & $0.000$\\
			$L$ factor & $0.027$ & $2$ & $0.014$ & $5.582$ & $0.005$\\
			$P$ factor & $1.031$ & $2$ & $0.516$ & $210.809$ & $0.000$\\
			$B$ factor & $0.089$ & $2$ & $0.045$ & $18.253$ & $0.000$\\
			$LP$ factors & $0.183$ & $4$ & $0.046$ & $18.695$ & $0.000$\\
			$PB$ factors & $0.124$ & $4$ & $0.031$ & $12.695$ & $0.000$\\
			Error & $0.294$ & $120$ & $0.002$ & & \\
			Total & $52.622$ & $135$ & $3.489$ & & 
		\end{tabular}
	\end{table}

	The results of the ANOVA III test for the Adience dataset are shown in Table \ref{table:ANOVAAdience}. The interactions where there are not significative differences have been omitted. First, the factor associated with the link function is analysed. The differences between the means are significative ($\text{p-value} = 0.005$). Secondly, there are significative differences concerning the learning rate factor too ($\text{p-value} = 0.000$). Also, there are differences for the means for the batch size factor. Significative interactions exist between two pairs of factors: link function and learning rate, and learning rate and batch size ($\text{p-value = 0.000}$ for both cases).

	A post-hoc multiple comparison test has been performed on the average WK obtained for this dataset too. Under the null hypothesis that the variance of the error of the dependent variable is the same between groups, an HSD Tukey's test has been done. The results showed that the best link function is the logit one, in this case, with $0.632438$ as mean WK value. However, there are not significative differences between this function and the complementary log-log ($\text{p-value}=0.110$). It has differences with the probit function though ($\text{p-value} = 0.003$). The complementary log-log link reported the best results after the logit function, having a mean WK of $0.611287$. However, it doesn't show significative differences with the probit function.
	The results have been obtained following a similar methodology with the different values of the learning parameter including the HSD Tukey's test and the learning parameter ranking. The test reported significative differences between $\eta=10^{-2}$ (mean value $0.733784$) and the rest of values of this parameter ($\alpha=0.005$), being this one the best value for this factor. The best value after $10^{-2}$ is $10^{-3}$, which have a mean WK value of $0.579889$. Lastly, the best value for the batch size parameter is $10$ (mean value $0.648700$) as it has significative differences with the rest of values. The second best value is $5$ (mean $0.605533$), but it has no significative differences with $15$ ($\text{p-value}= 0.194$).

@@

	The same experiments described in Section \ref{sect:experiments} were repeated using the cross-entropy instead of the QWK as loss function for the optimizer and the softmax function instead of the Proportional Odds Model for the output of the network.
	The results for both datasets are shown in Table \ref{table:NominalResults}.

@@

	As we did in Section \ref{sect:statisticalanalysis}, we must analyse the effect of the batch size (with values 5, 10 and 15) and the learning rate ($10^{-2}$, $10^{-3}$ and $10^{-4}$) factors have over the WK metric. So, we make an ANOVA II analysis for each dataset.

	In the Diabetic Retinopathy dataset, there are significative differences between the mean values of WK depending on the batch size and the initial value of the learning rate. Also, no significative interaction was detected between these factors.

@@

	A similar study was made for the Adience dataset. There are not significative differences for the mean value of WK w.r.t. the batch size factor. Also, there are no interactions between the batch size and the learning rate. The best results were obtained with $10^{-3}$ for the learning rate followed by $10^{-4}$, without significative differences between them.

	Finally, the conclusions of this study are exposed below:
	\begin{itemize}
		\item There are no interactions between both factors.
		\item Batch sizes 10 and 15 obtained the best results.
		\item Learning rate $10^{-4}$ is the best choice for both datasets as it the best for Diabetic Retinopathy, with significative differences, and it is the second for Adience, without significative differences with the best. 
	\end{itemize}

@@

	The proposed POM model has improved the performance of the deep network regarding Weighted Kappa compared to the model that uses the softmax function. This enhancement is more notable for the Diabetic Retinopathy dataset. Also, it reduces the chance that the model gets stuck when training with some parameter configurations. So, the most significant improvements of these link functions are the WK performance increase, the reduction of the number of parameters configurations that should be tried to find the best one and the prevention of the over-fitting and the stagnation.

>>>>

\renewcommand{\bibfont}{\footnotesize}

@@

	Ordinal classification problems are those where labels are ordered, and there are different inter-classes weights for each pair of classes. This kind of problem can be treated as a nominal classification problem, but when you do this, you are the discarding ordinal information. A better approach is to use specific methods that take into account this kind of information to improve the performance of the classification model. One way to use the ordinal information is to evaluate the model using an ordinal metric. Multiple metrics exist in the literature of machine learning and statistics \cite{cruz2014metrics, mehdiyev2016evaluating}. Kappa index is a well-known statistic coefficient defined by Cohen \cite{cohen1960coefficient} to measure inter-rater agreement on classifying elements into a set of categories. Later, Weighted Kappa (WK) is a modified version of the Kappa statistic calculated to allow assigning different weights to different levels of aggregation between two variables. weighted kappa loss was described in previous work~\cite{de2018weighted} and is a cost function that is based on the WK metric. These functions are indicated for problems where different inter-classes weights are assigned, and that work proved that using these functions improves the model performance and reduce the overfitting risk. This kind of problems is associated with ordinal data processing. Those weights are predefined and depend on the type of the chosen WK. In a linear penalization, the weight is proportional to the distance between the predicted and the real class. In the Quadratic Weighted Kappa (QWK) the penalization is proportional to the square of the distance. In this work, we will combine the QWK metric and the QWK loss.

@@

	This paper is organised as follows: in Section~\ref{sect:relatedwork}, we take a look at previous works related to this paper, in Section~\ref{sect:experiments} we describe the model, the experiments and the datasets used, in Section~\ref{sect:results} we present the results obtained and the statistical analysis and finally in Section~\ref{sect:conclusions} we expose the conclusions of this work.

@@

	\begin{figure}[!t]
		\centering
		\includegraphics[width=3in]{img/linkfunctions.png}
		\caption{Representation of link functions.}
		\label{fig:linkfunctions}
	\end{figure}

@@

	The model is evaluated with Quadratic Weighted Kappa metric (QWK)~\cite{ben2008comparison}. This evaluation measure gives a higher weight to the errors that are further from the correct class. Quadratic weighted kappa loss is considered as loss function for the optimizer as it gives better performance for ordinal classification problems.

@@

	\subsection{Diabetic Retinopathy}
	\label{sect:dr}
	Detailed results for the Diabetic Retinopathy dataset are presented in Table \ref{table:DRresults}.

@@

	The best mean QWK value was obtained with the complementary log-log link function using a batch size of 10 and a learning rate of $10^{-3}$.

	\subsection{Adience}
	\label{sect:adience}
	Results for the experiments made with the Adience dataset are shown in Table \ref{table:Adienceresults}.

@@

	The best mean QWK value was obtained with the probit link function using a batch size of 15 and a learning rate of $10^{-2}$.

@@

	We denote by $QWK_{i,j,k,l}(i=1, ..., 3; j = 1, ..., 3; k = 1, ..., 3)$ the value observed when the first factor is at the i-th level, the second at the j-th level and the third at k-th level. We assume that the three factors do not act independently and therefore there exists an interaction between each pair of them and between the three factors. In this case, the observations fit \ref{eq:anova3}.

	\begin{equation}
	\begin{aligned}
	\footnotesize
	QWK_{i,j,k,l} = & \mu + L_i + P_j + B_k + LP_{i,j} + LB_{i,k} \\& + PB_{j,k} + LPB_{i,j,k} + \epsilon_{i,j,k,l}
	\label{eq:anova3}
	\end{aligned}
	\end{equation}

@@

	The analysis of variance table containing the sum of squares, degrees of freedom, mean square, test statistics and significance level represent the initial study in a compact form, where non-significative factors and interactions have been removed ($\text{p-value} > 0.05$). These factors and interactions take part of the error component now.

	\begin{table}[!t]
		\caption{ANOVA III for the analysis of the main factors in the design of a Convolutional Ordinal Neural Network for the Retinopathy dataset.}
		\label{table:ANOVARetinopathy}
		\centering
		\small
		\begin{tabular}{l|ccccc}
			\multicolumn{6}{c}{Response variable QWK}\\\hline
			Source & S.S. & D.F. & M.S. & F-ratio & Sig.\\\hline
			Model & $37.860$ & $9$ & $4.207$ & $1562.840$ & $0.000$\\
			$L$ factor & $0.057$ & $2$ & $0.029$ & $10.646$ & $0.000$\\
			$P$ factor & $0.121$ & $2$ & $0.060$ & $22.468$ & $0.000$\\
			$LP$ factors & $0.057$ & $4$ & $0.014$ & $5.261$ & $0.001$\\
			Error & $0.339$ & $126$ & $0.003$ & & \\
			Total & $38.199$ & $135$ & & & 
		\end{tabular}
	\end{table}

	The results of the ANOVA III test for the Diabetic Retinopathy dataset are resumed in Table \ref{table:ANOVARetinopathy}. There are significative differences in the means of QWK depending on the link function and also depending on the learning rate for $\alpha=0.05$. Moreover, an interaction between the link function and the learning rate can be recognised ($\text{p-value} = 0.001$).

	As Table \ref{table:ANOVARetinopathy} showed that there exist significative differences between the means, we are now analysing those differences. A post-hoc multiple comparison test has been performed on the mean QWK obtained. An HSD Tukey's test was made under the null hypothesis that the variance of the error of the dependent variable is the same between groups. The best link function is the complementary log-log, though it doesn't have significative differences with the probit function ($\alpha=0.05$). There are differences between the complementary log-log and the probit function concerning the logit function.

@@

	\begin{table}[!t]
		\caption{ANOVA III for the analysis of the main factors in the design of a Convolutional Ordinal Neural Network for the Adience dataset.}
		\label{table:ANOVAAdience}
		\centering
		\small
		\begin{tabular}{l|ccccc}
			\multicolumn{6}{c}{Response variable QWK}\\\hline
			Source & S.S. & D.F. & M.S. & F-ratio & Sig.\\\hline
			Model & $52.328$ & $15$ & $3.489$ & $1426.170$ & $0.000$\\
			$L$ factor & $0.027$ & $2$ & $0.014$ & $5.582$ & $0.005$\\
			$P$ factor & $1.031$ & $2$ & $0.516$ & $210.809$ & $0.000$\\
			$B$ factor & $0.089$ & $2$ & $0.045$ & $18.253$ & $0.000$\\
			$LP$ factors & $0.183$ & $4$ & $0.046$ & $18.695$ & $0.000$\\
			$PB$ factors & $0.124$ & $4$ & $0.031$ & $12.695$ & $0.000$\\
			Error & $0.294$ & $120$ & $0.002$ & & \\
			Total & $52.622$ & $135$ & $3.489$ & & 
		\end{tabular}
	\end{table}

	The results of the ANOVA III test for the Adience dataset are shown in Table \ref{table:ANOVAAdience}. The interactions where there are not significative differences in QWK means have been omitted. First, the factor associated with the link function is analysed. The differences between the means are significative ($\text{p-value} = 0.005$). Secondly, there are significative differences concerning the learning rate factor too ($\text{p-value} = 0.000$). Also, there are differences for the means of QWK for the batch size factor. Significative interactions exist between two pairs of factors: link function and learning rate, and learning rate and batch size ($\text{p-value = 0.000}$ for both cases).

	A post-hoc multiple comparison test has been performed on the average QWK obtained for this dataset too. Under the null hypothesis that the variance of the error of the dependent variable is the same between groups, an HSD Tukey's test has been done. The results showed that the best link function is the logit one, in this case, with $0.60553$ as mean QWK value. However, there are not significative differences between this function and the complementary log-log ($\text{p-value}=0.110$). It has differences with the probit function though ($\text{p-value} = 0.003$). The complementary log-log link reported the best results after the logit function, having a mean QWK of $0.58738$. However, it doesn't show significative differences with the probit function.
	The results have been obtained following a similar methodology with the different values of the learning parameter including the HSD Tukey's test and the learning parameter ranking. The test reported significative differences between $\eta=10^{-2}$ (mean value $0.73378$) and the rest of values of this parameter ($\alpha=0.005$), being this one the best value for this factor. The best value after $10^{-2}$ is $10^{-3}$, which have a mean QWK value of $0.57989$. Lastly, the best value for the batch size parameter is $5$ (mean value $0.63243$) as it has significative differences with the rest of values. The second best value is $15$ (mean $0.61129$), but it has no significative differences with $10$ ($\text{p-value}= 0.194$).

	To sum up, the results showed that the best parameter configuration depends on the problem that is being solved. None of the optimal parameters is the same for Retinopathy and Adience datasets. These results highlight the importance of adjusting the hyper-parameters for each problem that is treated instead of trying to find an optimal configuration for all the datasets.

@@

	The same experiments described in Section \ref{sect:experiments} were repeated using the cross-entropy instead of the QWK as loss function for the optimizer and the softmax function instead of the Proportional Odds Model for the output of the network. The evaluation metric remains the same in order to be able to compare. The results for both datasets are shown in Table \ref{table:NominalResults}. There are some parameter configurations where the training process gets stagnated and a very low QWK is obtained. As we saw in Sections \ref{sect:dr} and \ref{sect:adience}, this problem is not present when using the ordinal method.

@@

	As we did in Section \ref{sect:statisticalanalysis}, we must analyse the effect of the batch size (with values 5, 10 and 15) and the learning rate ($10^{-2}$, $10^{-3}$ and $10^{-4}$) factors have over the QWK metric. So, we make an ANOVA II analysis for each dataset.

	In the Diabetic Retinopathy dataset, there are significative differences between the mean values of QWK depending on the batch size and the initial value of the learning rate. Also, no significative interaction was detected between these factors.

@@

	A similar study was made for the Adience dataset. There are not significative differences for the mean value of QWK w.r.t. the batch size factor. Also, there are no interactions between the batch size and the learning rate. The best results were obtained with $10^{-3}$ for the learning rate followed by $10^{-4}$, without significative differences between them.

	The conclusions obtained from the results are exposed below:
	\begin{itemize}
		\item There are no interactions between both factors.
		\item Batch sizes 10 and 15 obtained the best results.
		\item Learning rate $10^{-4}$ is the best choice for both datasets as it the best for Diabetic Retinopathy, with significative differences, and it is the second for Adience, without significative differences with the best. 
	\end{itemize}

	Finally, a comparison of the best QWK for each dataset for ordinal and nominal cases is shown in Table \ref{table:ComparisonNomOrd}. Also, it is compared with previous works. The proposed ordinal model outperforms all the other alternatives. The performance increase of POM over softmax reaches $17.07\%$ for Diabetic Retinopathy and $3.85\%$ for Adience dataset. The enhancement of the ordinal method for Retinopathy dataset is more notable than the rise for Adience dataset. It seems that the method proposed in this work offers a more significant improvement as the given problem complexity increases.

	\begin{table}[!t]
		\caption{Comparison between the best QWK results of nominal, ordinal and previous works.}
		\label{table:ComparisonNomOrd}
		\footnotesize
		\centering
		\begin{tabular}{lccc|cc}
			\multicolumn{6}{c}{Diabetic Retinopathy}\\
			\hline
			Method & BS & LR & LF & $\kappa_{val}$ & $\kappa_{test}$\\\hline
			Ord. proposed & 10 & $10^{-3}$ & c log-log & $0.58830$ & $0.58150$\\
			Nom. proposed & 10 & $10^{-4}$ & softmax & $0.49720$ & $0.49670$\\
			Best from \cite{de2018weighted} & 20 & $10^{-4}$ & softmax & $0.53700$ & -\\
			Best from \cite{nebot2016diabetic} & - & - & softmax & - & $0.55500$\\
			\hline\hline
			\multicolumn{6}{c}{Adience}\\
			\hline
			Ord. proposed & 15 & $10^{-2}$ & probit & $0.84254$ & $0.78666$\\
			Nom. proposed & 15 & $10^{-3}$ & softmax & $0.83090$ & $0.75750$\\
		\end{tabular}
	\end{table}

@@

	The proposed POM has improved the performance of the deep network regarding quadratic weighted kappa compared to the model that uses the softmax function. This enhancement is more notable for the Diabetic Retinopathy dataset. Also, it reduces the chance that the model gets stuck when training with some parameter configurations. So, the most significant improvements of these link functions are the QWK performance increase, the reduction of the number of parameters configurations that should be tried to find the best one and the prevention of the over-fitting and the stagnation.
